hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-01-08 inst/manuscript/paper.Rmd
add to model selection section
reknit

To widen the user application of these methods, we distribute two user-interfaces for **hatchR**. The first is a R package distributed via CRAN that allows users the most customizable application for these methods. The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes. These variable approaches are outlined in the package documentation on **hatchR**'s website. Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of **hatchR**'s functions in a graphical-user-interface. The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of **hatchR**'s R package. Below, we present the basic overview of the software and multiple case studies of how it may be applied.

@@

**hatchR** is meant to primarily be a tool for predicting phenology. In this sense, we mostly limit functionality to these applications and provide minimal data checking and plotting help. This decision is in part driven by the diversity of data types that users may import and the difficulty in addressing all those data types with respect to various data checks. In other words, we expect users to know their data better than we do and to check it accordingly. We do provide two basic data check functions discussed in the Checking Data section. Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualization from output, specifically in R. For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice. Finally, we provide brief summaries of general applications of **hatchR** below, but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.

@@

The effective value model framework is the basis for the phenological models in **hatchR**, both in the included `model_table` in the package (though `model_table` includes more complex models developed by @beacham1990), as well as for custom models users can fit with `fit_model()`. Specifically, `model_table` has includes parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).

@@

For users choosing to implement **hatchR** in program R, data can be imported from any format the user chooses, as long as users can eventually coerce their data into a `dataframe` or `tibble` format, in which each row represents a single record. For the Shiny application, users must have their data stored as a .csv (comma separated values) file for upload, which can easily be exported using datasheet software like Microsoft Excel or Google Sheets.

@@

**hatchR** is built assuming data will be analyzed as daily average temperatures. Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and **hatchR** includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to user's attention.

@@

talk about using nls for fit_model

model_table and fit_model

@@

Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package. We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach. We include parameterizations for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005. We demonstrate parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>.

@@

In the application of using effective value models **hatchR** and the user make some key assumptions. Particularly, numerous studies have indicated that stressful environmental conditions can cue fish to prematurely hatch or emerge from their environment. These include water quality like dissolved oxygen, pH, or salnity, pathogens, and even mechanical agitation (reviewed in @quinn2018 and @cowan2024). Moreover, while the **hatchR** provides point estimates for developmental phenology, spawning and developmental within populations of fishes both occur as distributions as opposed single events @mason1976 and we encourage users to either predict phenology using early, peak, and late frameworks (*e.g.*, 5th, 50th, and 95th quantiles) or with real or modeled distributions.

To date, the application of the effective value framework has largely focused using species-specific models to predict phenology in wild environments (see @adelfio2024, @austin2019). However, we emphasize that the statistical relationships at the basis of these models are fundamentally reaction norms, meaning that both the intercept and slope of the responses of fishes to temperature are reflective of family- or population-level genetic variation, genetic x environment responses, or may be indicative of phylogenetic differences among species @west-eberhard2003. For instance, while @sparks2017 did not find differences in the rates of development between the focal populations of their study, they did observe family-level genetic x environment interactions across different thermal regimes. Similarly, when they reparameterized their models using fish from western Alaska, they found differences in both the slope and intercept of the model relative to the original parameterization from @beacham1990, notably derived from multiple Canadian stocks. The western Alaskan fish generally developed more slowly than their southerly counterparts, consistent with cogradient variation @conover2009, @sparks2022. In this sense, we encourage users not to only think about the end points of these models (days to development) but also how the statistical relationships they are premised on inform our understanding of micro- and and macro-evolutionary processes.

talk about how these models represent point estimates and that emergence and hatch will take the form of a distribution

talk about how temperatures may not always reflect redd temperture

talk about how other factors like oxygen or predator cues may change phenology

talk about how these models could be extended for other development phases or to organisms that fit similar developmental patterns

talk about how advanced users could fit their own models and provide their custom expression to predict_phenology for a really customized option.

>>>>

To widen the user application of these methods, we distribute two user-interfaces for **hatchR**. The first is a R package (@r:ther) distributed via CRAN that allows users the most customizable application for these methods. The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes. These variable approaches are outlined in the package documentation on **hatchR**'s website. Alternatively, we also distribute a Shiny application (@chang2024) in the form of an HTML-based web tool to interact with many of **hatchR**'s functions in a graphical-user-interface. The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of **hatchR**'s R package. Below, we present the basic overview of the software and multiple case studies of how it may be applied.

@@

**hatchR** is meant to primarily be a tool for predicting phenology. In this sense, we limit functionality to these applications and provide minimal data checking and plotting help. This decision is in part driven by the diversity of data types that users may import and the difficulty in addressing all those data types with respect to various data checks. In other words, we expect users to know their data better than we do and to check it accordingly. We do provide two basic data check functions discussed in the Checking Data section. Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualizations from output, specifically in R. For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice. Finally, we provide brief summaries of general applications of **hatchR** below, but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.

@@

The effective value model framework is the basis for the phenological models in **hatchR**, both in the included `model_table` in the package (though `model_table` includes more complex models developed by @beacham1990), as well as for custom models users can fit with `fit_model()`. Specifically, `model_table` includes parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).

@@

For users choosing to implement **hatchR** in R, data can be imported from any format the user chooses, as long as users can eventually coerce their data into a `dataframe` or `tibble` format, in which each row represents a single record. For the Shiny application, users must have their data stored as a .csv (comma separated values) file for upload, which can easily be exported using datasheet software like Microsoft Excel or Google Sheets.

@@

**hatchR** is built assuming data will be analyzed as daily average temperatures. Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and **hatchR** includes functionality to summarize those data into a format that is usable (only in R, it must be summarized for the Shiny app), as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to user's attention.

@@

Users can either use Salmonid models from `model_table` included in the package and the Shiny app. As discussed, these models are included because parameterizations already existed in the literature. However, these parameterizations are limited to Pacific Salmon and Bull Trout. To widen the application of the effective value approach, we include a `fit_model()` function and is species agnostic (as long as develop generally follows the power law.

The function `fit_model()` uses data in which average incubation temperature (°C) and days to phenological event (as two vectors) are the inputs and estimates parameter coeficients for *log~e~a* and *b* using `stats::nls()`

**BRYAN TO FINISH THIS PARAGRAPH OFF HERE**

@@

Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package. We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for fishes beyond the scope of the original effective value approach. These parameterizations are for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005. We demonstrate parameterization for just Smallmouth Bass here for concision, but the full code set for all species is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>.

@@

In the application of using effective value models **hatchR** and the user make some key assumptions. Particularly, numerous studies have indicated that stressful environmental conditions can cue fish to prematurely hatch or emerge from their environment. These include water quality like dissolved oxygen, pH, or salinity, pathogens, and even mechanical agitation (reviewed in @quinn2018 and @cowan2024). Moreover, while the **hatchR** provides point estimates for developmental phenology, spawning and developmental within populations of fishes both occur as distributions as opposed single events @mason1976 and we encourage users to either predict phenology using early, peak, and late frameworks (*e.g.*, 5th, 50th, and 95th quantiles) or with real or modeled distributions. Another factor that could bias estimates from these models is that temperatures used from sensors don't reflect the geomoprhic properties in the environment where eggs are developing such that they may be too cold or warm @geist2002.

To date, the application of the effective value framework has largely focused using species-specific models to predict phenology in wild environments (see @adelfio2024, @austin2019). However, we emphasize that the statistical relationships at the basis of these models are fundamentally reaction norms, meaning that both the intercept and slope of the responses of fishes to temperature are reflective of family- or population-level genetic variation, genetic x environment responses, or may be indicative of phylogenetic differences among species (@west-eberhard2003). For instance, while @sparks2017 did not find differences in the rates of development between the focal populations of their study, they did observe family-level genetic x environment interactions across different thermal regimes. Similarly, when they reparameterized their models using fish from western Alaska, they found differences in both the slope and intercept of the model relative to the original parameterization from @beacham1990, notably derived from multiple Canadian stocks. The western Alaskan fish generally developed more slowly than their southerly counterparts, consistent with cogradient variation (@conover2009, @sparks2022). In this sense, we encourage users not to only think about the end points of these models (days to development) but also how the statistical relationships they are premised on inform our understanding of micro- and and macro-evolutionary processes.

**BRYAN check the below**

The models presented in **hatchR** can be further customized in multiple ways beyond the use-cases provided above. For instance, while our models are fit to predict hatch or emergence timing, they could be used to predict other developmental stages prior to the initation of external feeding like eye-up or other embryological developmental stages (@velsen1980) or **Some example from a non-salmonid, maybe initiation of pelagic-larval or riverine current dispersal**. Additionally, while `fit_model()` relies on non-linear modeling platform, `predict_phenology()` only requires users to pass a model expression, so if users preferred a different model fit than the non-linear option provided (as long as it integrates daily temperature), they could pass a different expression into `predict_phenology()` for additional customization. Finally, while **hatchR** was designed specifically for fishes, we expect other poikilothermic organisms such as reptiles, amphibians, or invertebrates would develop in accordance with the power law and these developmental models could be extended to other organisms beyond fishes.
